---
title: "MethylationHW"
author: "Milo Gittings"
date: "2026-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("/home1/gittings/490_cluster/analysis_data"))
```

```{r}
library(TCGAbiolinks)
library(sesame)
library(sesameData)
library(limma)
library(ggplot2)
```

```{r}
query <- GDCquery(
  project = "TCGA-BRCA",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation 450",
  data.format = "Dataframe"
)

# GDCdownload(query, method = "api", files.per.chunk = 10)
methylation450 <- GDCprepare(query)
```

```{r}
methylation_clinical <- as.data.frame(methylation450@colData)

betas <- as.data.frame(methylation450@assays@data@listData) #takes a longer time

cpg_sites <- as.data.frame(methylation450@rowRanges@elementMetadata)

# Masking
column_mask <- ifelse(colnames(methylation_clinical) %in% c('treatments', 'primary_site', 'disease_type','sites_of_involvement'), F, T)

methylation_clinical <- methylation_clinical[,column_mask]

write.csv(methylation_clinical, 'brca_methylation_clinical.csv')

site_mask <- !grepl('-', cpg_sites$gene) & !grepl(';', cpg_sites$gene) & !is.na(cpg_sites$gene) & complete.cases(betas)

betas <- betas[site_mask,]
cpg_sites <- cpg_sites[site_mask,]
write.csv(betas, 'brca_methylation_betas.csv')
write.csv(cpg_sites, 'brca_cpg_sites.csv')
```

```{r}
meth_mask <- methylation_clinical$definition %in% c("Metastatic", "Primary solid Tumor")
methylation_clinical <- methylation_clinical[meth_mask, ]
betas_clean <- betas[, meth_mask]
methylation_clinical$is_metastatic <- methylation_clinical$definition == "Metastatic"
non_metastatic_mask <- methylation_clinical$metastatic_status == "Non-Metastatic"
mval <- t(apply(betas_clean, 1, function(x) log2(x/(1-x))))
design <- model.matrix(~is_metastatic, data = methylation_clinical)
fit <- lmFit(mval, design)
fit2 <- eBayes(fit)
```

```{r}
dat <- data.frame(foldchange = fit[["coefficients"]][,2], logPvalue =  -log10(p.adjust(fit2[["p.value"]][,2],method='BY')), geneName = cpg_sites$gene)
dat$threshold <- as.factor(abs(dat$foldchange) < 1)
colnames(betas_clean) <- gsub("\\.", "-", colnames(betas_clean))
cols <- c("TRUE" = "grey", "FALSE" = "blue")
ggplot(data=dat, aes(x=foldchange, y = logPvalue)) +
  geom_point(alpha=.2, size=0.6) +
  scale_colour_manual(values = cols) +
  geom_vline(xintercept = 1, colour="darkred", linetype="dashed") + 
  geom_vline(xintercept = - 1, colour="darkred", linetype="dashed") +
  geom_hline(yintercept = 2, colour = "darkred", linetype="dashed") +
  theme(legend.position="none") +
  xlab("Fold Change") +
  ylab("-log10 p value") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
write.csv(dat, "methylation_results_metastasis.csv")
```
This volcano plot is showing that most of the CpG sites don't have significant differential methylation between metastatic and non-metastatic patients, because the majority are below the significance threshold. Between the data in the upper left and upper right quadrants (significantly hypo and hyper methylated sites), there is shown to be more hypomethylation than hypermethylation, which suggests that genes for metastatic patients are more commonly under-methylated in comparison to over-methylatd. Therefore showing that these hypomethylated genes could play a role in the metastasis of breast cancer




